Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2016-11-10 paper/algorithms_v3.tex
add comments from karen on version 2 (not on abstract or intro yet); create v3 files to start incorporating these comments

Just as error estimates can be used to guide mesh-refinement~\cite{BecRann01,VendDarm00,PrudOden99,BecVex05,AinsOden11}, the error estimate \cref{eq:finErrExp} can be localized to give elemental contributions and used to guide the division of the domain for a mixed-fidelity model. The error estimate can be calculated again, using the mixed-fidelity model as the lower-fidelity model. This process can be repeated, successively increasing the proportion of the domain in which the high-fidelity model is used, until some threshold is reached. \Cref{alg:refSeries} describes this approach. Note that the QoI error adjoint problem \cref{eq:superAdjEq} involves linearization about $\Psi_{HF}$, which is not available, so in the case of a nonlinear goal-oriented inverse problem, the QoI error adjoint problem is approximated by linearizing about $\Psi_{LF}$ instead.
%
\alglanguage{pseudocode}
\begin{algorithm}[h!]
\small
\caption{An algorithm to adaptively build a mixed-fidelity model for low error in the QoI.}
\label{alg:refSeries}
\begin{algorithmic}[1]
\State{Define maximum acceptable absolute relative QoI error \texttt{errTol}}
\State{Define maximum number of adaptive iterations \texttt{maxIter}}
\Procedure{$\texttt{BuildMF}$}{HF model, LF model, \texttt{errTol}, \texttt{maxIter}}
	\State{Let the model MF$_0$ be the LF model applied everywhere in the domain.}
	\State{$i\gets0$}
	\State{Solve for stationary point $\Psi_{MF_0}$ of augmented Lagrangian $\mathcal{M}_{MF_0}$}
	\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_0}$, for \par\hskip\algorithmicindent supplementary adjoint $\Lambda_0$ (see \cref{eq:superAdjEq})}
	\State{Compute QoI error estimate

@@

\Cref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh-refinement, though in this case the method described in~\cite{BecVex05} would likely be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.

@@

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. For ease of implementation, we can choose these basis functions to be the same as those spanning the space in which we solve for the primary, auxiliary, and supplementary adjoint vatriables. Near the interfaces between the low-fidelity and high-fidelity regions, basis functions may have their support divided between the two regions and thus have a nonzero error contribution. The basis functions with the largest error contributions are flagged, and the elements in their support are added to the high-fidelity subdomain. Basis functions whose support lie completely in the high-fidelity region will have a zero contribution and the elements in their support are never flagged for refinement again.

>>>>

Just as error estimates can be used to guide mesh refinement~\cite{BecRann01,VendDarm00,PrudOden99,BecVex05,AinsOden11}, the error estimate \cref{eq:finErrExp} can be localized to give elemental contributions and used to guide physics-based refinement over the domain to create a mixed-fidelity model. The error estimate can be calculated again, using the mixed-fidelity model as the lower-fidelity model. This process can be repeated, successively increasing the proportion of the domain in which the high-fidelity model is used, until some threshold is reached. \Cref{alg:refSeries} describes this approach. Note that the QoI error adjoint problem \cref{eq:superAdjEq} involves linearization about $\Psi_{HF}$, which is not available, so in the case of a nonlinear goal-oriented inverse problem, the QoI error adjoint problem is approximated by linearizing about $\Psi_{LF}$ instead.
%
\alglanguage{pseudocode}
\begin{algorithm}[h!]
\small
\caption{An algorithm to adaptively build a mixed-fidelity model for low error in the QoI.}
\label{alg:refSeries}
\begin{algorithmic}[1]
\State{Define maximum acceptable absolute relative QoI error \texttt{errTol}}
\State{Define maximum number of adaptive iterations \texttt{maxIter}}
\Procedure{$\texttt{BuildMixedFidelity}$}{HF model, LF model, \texttt{errTol}, \texttt{maxIter}}
	\State{Let the model MF$_0$ be the LF model applied everywhere in the domain.}
	\State{$i\gets0$}
	\State{Solve for stationary point $\Psi_{MF_0}$ of augmented Lagrangian $\mathcal{M}_{MF_0}$}
	\State{Solve QoI error adjoint equation, linearized about $\Psi_{MF_0}$, for \par\hskip\algorithmicindent supplementary adjoint $\Lambda_0$ (see \cref{eq:superAdjEq})}
	\State{Compute QoI error estimate

@@

\Cref{alg:refSeries} is applicable to a large class of models. The lower-fidelity model could, for example, be a simplified model including fewer physical phenomena, be a reduced-order model, or have a reduced parameter space. The two models could also correspond to two levels of mesh refinement, though in this case the method described in~\cite{BecVex05} would likely be more efficient, since interpolation could be used to estimate $\Psi_{HF}-\Psi_{LF}$ instead. The derived error estimate is not applicable to all models, however. The two models have to be expressed in a weak form, so this cannot be applied to, for example, a model of chemical reactions using kinetic Monte Carlo. We need some degree of compatibility between the two models; namely, we assume that $\Psi_{LF}$ will be in a space admissible to $\mathcal{M}'_{HF,\Psi}$, and that the QoI functional $I$ is applicable to both $(q_{HF},u_{HF})$ and $(q_{LF},u_{LF})$.

@@

We instead use the alternative method described in \cite{vanOpstaletal15}, decomposing the error estimate into contributions from locally supported basis functions rather than elements. For ease of implementation, we can choose these basis functions to be the same as those spanning the space in which we solve for the primary, auxiliary, and supplementary adjoint variables. Near the interfaces between the low-fidelity and high-fidelity regions, basis functions may have their support divided between the two regions and thus have a nonzero error contribution. The basis functions with the largest error contributions are flagged, and the elements in their support are added to the high-fidelity subdomain. Basis functions whose support lie completely in the high-fidelity region will have a zero contribution and the elements in their support are never flagged for refinement again.

### 2016-11-10 paper/mathematical_formulation_v2_kw.tex
add comments from karen on version 2 (not on abstract or intro yet); create v3 files to start incorporating these comments

\section{Mathematical Formulation}\label{sec:form}
%
We first introduce the goal-oriented inverse problem formally. Then we derive a rigorous a posteriori error estimate for the error induced in the QoI due to the use of mixed/lower-fidelity models.

%------------------------------------------------------------------------------%
\subsection{Problem Setup}  \label{sec:setup}
%------------------------------------------------------------------------------%
%
Consider a model for which the Galerkin formulation of the weak form is written as
\begin{equation}
a(u,q)(\phi)=\ell(q)(\phi),\quad\forall\phi\in U,
\label{eq:weakForm}
\end{equation}
where $u\in U$ is the state, $q\in Q$ are the unknown parameters, $\phi$ is a test function, and the state space $U$ and parameter space $Q$ are Hilbert spaces. The form $a$ and functional $\ell$ are linear with respect to the arguments in the second pair of parentheses (in \cref{eq:weakForm}, they are linear with respect to $\phi$). Further, we define an observation operator $C:U\to\Reals^{n_d}$ that maps the state to $n_d$ predicted observations. The actual observations (data) are denoted by $d\in\R^{n_d}$.

The unknown parameter $q$ can be inferred by minimizing the difference between the predicted and actual observations, leading to an inverse problem. Such inverse problems are typically ill-posed, since the observations are sparse, and insufficiently informative to uniquely determine the parameters. To make the inverse problem well-posed, a regularization, denoted by $R(q)$, is used to inject prior information or beliefs about the parameters into the formulation. The regularized inverse problem can thus be written as a constrained optimization problem,
%
\begin{subequations}
\label{eq:invOpt}
\begin{align}
\min\limits_{q,u} & \quad J(q,u)=\frac{1}{2}\|d-C(u)\|_2^2 + R(q), \label{eq:invOpt_obj} \\
\textrm{s.t. }& \quad a(u,q)(\phi)=\ell(q)(\phi),\quad\forall\phi\in U. \label{eq:invOpt_cons}
\end{align}
\end{subequations}
%
Thus, we aim to minimize the cost function $J$, which includes the mismatch between predicted and actual observations and a regularization term $R(q)$, subject to the state $u$ and parameters $q$ satisfying the model given by \cref{eq:weakForm}, which appears as a constraint in \cref{eq:invOpt_cons}. The constraints \cref{eq:invOpt_cons} are typically models of physical processes or systems. A given physical system can be described to varying degrees of fidelity using different models. This naturally introduces a hierarchy of models and a tradeoff between fidelity and computational expense; how one manages this tradeoff depends on what aspect of the solution one is interested in.

In the case of a goal-oriented inverse problem, the ultimate purpose of inferring the unknown parameters is to calculate some Quantity of Interest (QoI). Assuming a single scalar QoI, we denote this QoI by $I(q,u)$, where $I:Q\times U\to\R$ is a functional that maps the parameters and state to our QoI. Since our ultimate goal is to compute this QoI, the particular tradeoff we consider is in the fidelity and resultant computational expense of the model we use, versus the error in the QoI. We thus seek to introduce the QoI functional $I$ into the inverse problem formulation by introducing auxilary variables and additional adjoint equations. Then we use this formulation to derive an a posteriori error estimate for $I$, where the errors considered are those which occur due to the use of difference models in the constraint \cref{eq:invOpt_cons}.
%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of models, consider the QoI calculated from inferring the parameters with the highest-fidelity model; we take this QoI to be the value with which we compare other QoI estimates. In this section we derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model. 
%
\begin{theorem}
\label{thm:error_estimate}
Consider the inverse problem described by the constrained optimization problem \cref{eq:invOpt}. Let the form $a:U \times U \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q\to\Reals$ be differentiable with respect to the parameter $q$, and the functional $I:Q\times U\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

@@

Consider two models with which parameters can be inferred: a high-fidelity model and a lower-fidelity model. Let the high-fidelity (HF) and low-fidelity (LF) models, and their corresponding variables, Lagrangians, and augmented Lagrangians, be distinguished by the subscripts $_{HF}$ and $_{LF}$, respectively. In particular, let $\Psi_{HF}$ and $\Psi_{LF}$ denote the stationary points of the high- and low- fidelity augmented Lagrangians $\mathcal{M}_{HF}$ and $\mathcal{M}_{LF}$, respectively. Consider the adjoint problem
%
\begin{equation}
\label{eq:superAdjEq}
\mathcal{M}'_{HF,\Psi}(\Lambda;\Psi_{HF})(\Phi)=\mathcal{Q}(\Phi)=\mathcal M'_{HF}(\Psi_{LF})(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2,
\end{equation}
%
for the supplementary adjoint $\Lambda$, where $\Phi$ is a test function. Then, the error in the Quantity of Interest $I$ is given by,
%
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF})+\mathcal{R}(e^3).
\end{multline}
%
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$.
\end{theorem}
%
\begin{proof}
%
Observe that,
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

Extending the property in \cref{eq:MeqI} to the augmented Lagrangians for the high and low fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Applying Proposition 3 from~\cite{BecVex05} for the difference $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$,
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{equation}
Combining \cref{eq:repIwithM} and \cref{eq:beckvex} we obtain
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{multline}

>>>>

\section{Mathematical Formulation}\label{sec:form}
%
We first introduce the goal-oriented inverse problem. Then we derive a rigorous a posteriori error estimate for the error induced in the QoI due to the use of mixed-fidelity and/or lower-fidelity models.

%------------------------------------------------------------------------------%
\subsection{Inverse Problem Formulation}  \label{sec:setup}
%------------------------------------------------------------------------------%
%
Consider a model for which the Galerkin formulation of the weak form is written as
\begin{equation}
a(u,q)(\phi)=\ell(q)(\phi),\quad\forall\phi\in U,
\label{eq:weakForm}
\end{equation}
where $u\in U$ is the state, $q\in Q$ are the unknown parameters, $\phi$ is a test function, and the state space $U$ and parameter space $Q$ are Hilbert spaces. The form $a$ and functional $\ell$ are linear with respect to the arguments in the second pair of parentheses (in \cref{eq:weakForm}, they are linear with respect to $\phi$). Further, we define an observation operator $C:U\to\Reals^{n_d}$ that maps the state to $n_d$ predicted observations. The actual observations (data) are denoted by $d\in\R^{n_d}$.

The unknown parameter $q$ can be inferred by minimizing the difference between the predicted and actual observations, leading to an inverse problem. In our setting, such inverse problems are typically ill-posed, since the observations are sparse, and insufficiently informative to uniquely determine the parameters. To make the inverse problem well-posed, a regularization, denoted by $R(q)$, is used to inject prior information or beliefs about the parameters into the formulation. The regularized inverse problem can thus be written as a constrained optimization problem,
%
\begin{subequations}
\label{eq:invOpt}
\begin{align}
\min\limits_{q,u} & \quad J(q,u)=\frac{1}{2}\|d-C(u)\|_2^2 + R(q), \label{eq:invOpt_obj} \\
\textrm{s.t. }& \quad a(u,q)(\phi)=\ell(q)(\phi),\quad\forall\phi\in U. \label{eq:invOpt_cons}
\end{align}
\end{subequations}
%
Thus, we aim to minimize the cost function $J$, which includes the mismatch between predicted and actual observations and a regularization term $R(q)$, subject to the state $u$ and parameters $q$ satisfying the model given by \cref{eq:weakForm}, which appears as a constraint in \cref{eq:invOpt_cons}. The constraints \cref{eq:invOpt_cons} are typically models of physical processes or systems. A given physical system can be described to varying degrees of fidelity using different models. This naturally introduces a hierarchy of models and a tradeoff between fidelity and computational expense; how one manages this tradeoff depends on what aspect of the solution is of interest.

We consider the case of a {\em goal-oriented inverse problem}, where the ultimate purpose of inferring the unknown parameters is to calculate some quantity of interest (QoI). We denote a scalar QoI by $I(q,u)$, where $I:Q\times U\to\R$ is a functional that maps the parameters and state to our QoI. We then consider the tradeoff between the error in the QoI and the fidelity (and corresponding computational expense) of the model we use. In the following, we introduce the QoI functional $I$ into the inverse problem formulation by introducing auxiliary variables and additional adjoint equations. We then use this formulation to derive an a posteriori error estimate for $I$, where the errors considered are those due to the use of different multi-fidelity models in the constraint \cref{eq:invOpt_cons}.
%
%------------------------------------------------------------------------------%
\subsection[Error Estimate for a Goal-Oriented Inverse Problem]{Error Estimate for a Goal-Oriented Inverse Problem}  \label{sec:deriv}
%------------------------------------------------------------------------------%
%
For a given hierarchy of models, consider the QoI calculated from inferring the parameters with the highest-fidelity model; we take this QoI to be the value with which we compare other QoI estimates. In this section we derive an a posteriori estimate for the error in the QoI from inferring the parameters with a lower-fidelity model.
%
\begin{theorem}
\label{thm:error_estimate}
Consider the inverse problem described by the constrained optimization problem \cref{eq:invOpt}. Let the form $a:U \times U \to \Reals$ be three times continuously differentiable with respect to the state $u$ and parameters $q$. Let the observation operator $C:U\to\Reals^{n_d}$ be three times continuously differentiable with respect to the state $u$. Also, let the regularization operator $R:Q\to\Reals$ be differentiable with respect to the parameter $q$, and the functional $I:Q\times U\to\Reals$ be differentiable with respect to the state $u$ and parameter $q$.

@@

Consider two models with which parameters can be inferred: a high-fidelity model and a lower-fidelity model. Let the high-fidelity (HF) and low-fidelity (LF) models, and their corresponding variables, Lagrangians, and augmented Lagrangians, be distinguished by the subscripts $_{HF}$ and $_{LF}$, respectively. In particular, let $\Psi_{HF}$ and $\Psi_{LF}$ denote the stationary points of the high- and low- fidelity augmented Lagrangians $\mathcal{M}_{HF}$ and $\mathcal{M}_{LF}$, respectively. Consider the adjoint problem
%
\begin{equation}
\label{eq:superAdjEq}
\mathcal{M}'_{HF,\Psi}(\Lambda;\Psi_{HF})(\Phi)=\mathcal{Q}(\Phi)=\mathcal M'_{HF}(\Psi_{LF})(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2,
\end{equation}
% KW: notation in the equation is \mathcal{M}'_{HF,\Psi}(\Lambda;\Psi_{HF})= M'_{HF}(\Psi_{LF})(\Phi)
% I don't understand what the difference in notation means. Why drop the second subscript on the M'? and why the change in arguments? And neither instance is consistent with how the ' notation was defined above for the Lagrangian.
% You can't just introduce new notation and hope that the reader guesses what you mean. Everything must be defined! Every time you change an argument form, every time you change the form of the subscript, every time you change anything, ...
% KW: where is \mathcal{Q} defined?
%
for the supplementary adjoint $\Lambda$, where $\Phi$ is a test function. Then, the error in the Quantity of Interest $I$ is given by,
%
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF})+\mathcal{R}(e^3).
\end{multline}
%
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$.
\end{theorem}
%
\begin{proof}
%
Observe that
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

Extending the property in \cref{eq:MeqI} to the augmented Lagrangians for the high- and low-fidelity models, we have,
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Applying Proposition 3 from~\cite{BecVex05} 
% KW: what is Prop 3? is it something we need to state/summarize here for completeness? maybe okay as is, just seems a little mysterious.
for the difference $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$,
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{equation}
Combining \cref{eq:repIwithM} and \cref{eq:beckvex} we obtain
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{multline}
